### 2019-06-15 R/paper 02 Scale-jump analysis of historical retrospective patterns.Rmd
Update 10 june 2019

---
output: 
  word_document:
    reference_docx: ../report_template_v1.5.dotx
bibliography: paper_02_scalejumps.bib
csl: ices-journal-of-marine-science.csl
link-citations: yes
---

@@

fig_nums(name    = "retrometrics", level = 1, display = FALSE, 
         caption = " Retrospective metrics for 9 species derived from historical assessment records. Metrics ab (average bias) and asd (average variation) from Jonsson & Hjorleifsson 2000, rho from Mohn 1999 and sigma from Ralston et al 2011. In red: calculated values based on the whole time-series of historical data. In blue: based on retrospective peels of 5 years. Metrics ab, asd and sigma on a log scale, rho on a ratio scale.")

fig_nums(name    = "retrometrics2", level = 2, display = FALSE, 
         caption = "Summary of retrospective metrics for 9 species derived from historical assessment records. Metrics ab (average bias) and asd (average variation) from Jonsson & Hjorleifsson 2000, rho from Mohn 1999 and sigma from Ralston et al 2011. In red: calculated values based on the whole time-series of historical data. In blue: based on retrospective peels of 5 years. Metrics ab, asd and sigma on a log scale, rho on a ratio scale.")

@@

tab_nums(name    = "scalejumps", level = 1, display = FALSE, 
         caption = "Percentage assessments with scale jumps of more than 10% expressed by decade. Between brackets: number of assessments within the decade.")

fig_nums(name    = "scalecomparisons", level = 1, display = FALSE, 
         caption = "Example of pairwise assessment comparisons for Northeast Atlantic mackerel  over a period of 10 years in common. The dotted lines indicate the average values of the different assessments which are used to calculate scale differences between assessments.")

fig_nums(name    = "scalejumps", level = 2, display = FALSE, 
         caption = "Retrospective scale difference (sigma) for 9 species derived from historical assessment records. Results from pairwise comparisons over periods of 10 years. Large dots  refer to years when the scale difference is larger than 10%. Colours indicate the decades. Black triangles indicate benchmark assessment years. The bottom part of the panels shows the assessment methods being deployed, the upper part of the panels the stock definitions.")

fig_nums(name    = "metricsinpeels", level = 2, display = FALSE, 
         caption = "Comparison of retrospective uncertainty metrics from historical assessment records, calculated in peels, normalized between 0 and 1. Asd from Jonsson & Hjorleifsson 2000, rhoabs as absolute value of rho from Mohn 1999, sigma from Ralston et al 2011, percjumps as the percentage of jumps (this paper) and phiabs as the absolute value of phi (this paper).")

# section 4 -----------------------------------------------------

@@

**Tables**

*`r tab_nums("scalejumps")`*

```{r}

# print table
d.jumps.bydecade %>% 
  filter(variable == "stocksize") %>% 
  ungroup() %>% 
  select(stock, decade, ratio3) %>% 
  spread(key=decade, value=ratio3) %>% 
  pandoc.table(.,
               style = "simple",
               split.tables=400, 
               justify = "left",
               missing=".",
               round=c(0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0))

@@

<!-- ################################################################################## -->
<!-- Plot different methods of creating historical retrospectives --------------------- -->
<!-- ################################################################################## -->

```{r echo=FALSE, fig.align="center", fig.asp=0.7, message=FALSE, warning=FALSE}

knitr::include_graphics("fig1.2_retromethods.jpg")

*`r fig_nums("differentmethods")`*

@@

myvar         <- "stocksize"
myspecies     <- "her"
firstyear     <- 2008
firstplotyear <- 2000
mynyears      <- 5
myremoveyears <- 0
mythreshold   <- 0.1
mytitle       <- paste0(toupper(myspecies), ", ", myvar, ", number of years: ",mynyears, 
                        ", removed years: ", myremoveyears)

@@

@@

# threshold <- 0.10
mynyears      <- 10
myremoveyears <- 0
mythreshold   <- 0.1
myvar         <- "stocksize"
mytitle       <- paste0(myvar, ", number of years: ",mynyears, ", threshold: ", 
                        comma_format(digits=1, suffix="%")(100*mythreshold))

@@

@@

myremoveyears <- 0
myvar         <- "stocksize"
myspecies     <- "cod"
mytitle       <- paste0(toupper(myspecies), ", ", myvar, ", removed years: ", myremoveyears)

@@

```{r metricsatstockchange, echo=FALSE, fig.asp=1.0, fig.align="center", message=FALSE, warning=FALSE}

@@

@@

  geom_line(aes(colour=label), size=0.5) +

  geom_dl(aes(label  = stockkeylabelold, colour = label),
          method = list(dl.combine("first.bumpup"), cex = 0.8, hjust=0)) +

@@

@@

```{r jumpratio, eval=FALSE, fig.align="center", fig.asp=0.5, message=FALSE, warning=FALSE, include=FALSE}

@@

```{r metricsinpeels, echo=FALSE, fig.asp=1.0, fig.align="center", message=FALSE, warning=FALSE}

@@

>>>>

---
output: 
  word_document:
    reference_docx: ../report_template_v1.5.dotx
---

@@

@@

@@

@@

@@

```{r, echo=FALSE, fig.asp=1.4, fig.align="center", message=FALSE, warning=FALSE}

fig_nums(name    = "pairwisecomparisons", level = 1, display = FALSE, 
         caption = "Pairwise assessment comparisons for the five most recent assessment pairs over a period of 5  years in common. The dotted lines indicate the average values of the different assessments which are used to calculate scale differences between assessments.")

myvar         <- "stocksize"
# myvar         <- "fishingpressure"
myspecies     <- c("cod", "mac", "hom", "had", "her","hke","ple","sol","whb")
firstyear     <- 2013
firstplotyear <- 2000
lastplotyear  <- 2020
mynyears      <- 5
myremoveyears <- 0
mythreshold   <- 0.2
mytitle       <- paste0(myvar, ", number of years: ",mynyears, 
                        ", removed years: ", myremoveyears)

# selection of pairwise means
d.pairwise.means <-

  res$d.pairwise %>% 
  # ungroup() %>% 
  # distinct(removelastyears, numberofyears, threshold) %>% 
  # View()
  mutate(species = substr(stockkeylabelold,1,3)) %>%

  filter(removelastyears==myremoveyears, numberofyears==mynyears) %>% 
  filter(variable == myvar) %>% 
  filter(species %in% myspecies) %>% 
  filter(as.numeric(substr(assessmentpair,1,4)) >= firstyear) %>%

  # calculate means per period
  group_by(stock, species, assessmentpair, variable) %>% 
  mutate(   firstyear      = min(year),
            lastyear       = max(year)) %>% 
  group_by(stock, species, assessmentpair, variable, assessmentyear, label, firstyear, lastyear) %>% 
  summarise(value = mean(value, na.rm=TRUE)) %>% 
  gather(key=tmp, value=year, firstyear:lastyear) %>% 
  arrange(stock, species, assessmentpair, variable, assessmentyear, label)

# selection of pairwise long timeseries
d.pairwise.roughly <-

  res$d.pairwise.roughly %>% 
  mutate(species = substr(stockkeylabelold,1,3)) %>% 
  filter(removelastyears==myremoveyears, numberofyears==mynyears) %>% 
  filter(variable == myvar) %>% 
  filter(species %in% myspecies) %>% 
  filter(as.numeric(substr(assessmentpair,1,4)) >= firstyear)

# selection of pairwise phi texts
d.pcm <-
  res$d.pcm %>% 
  mutate(species = substr(stock,1,3)) %>% 
  filter(variable == myvar) %>% 
  filter(species %in% myspecies) %>% 
  filter(as.numeric(substr(assessmentpair,1,4)) >= firstyear) %>% 
  filter(numberofyears == mynyears) %>% 
  filter(removelastyears == myremoveyears) %>% 
  filter(threshold == mythreshold)

# start plot
res$d.pairwise %>%

  mutate(species = substr(stockkeylabelold,1,3)) %>%

  filter(removelastyears==myremoveyears, numberofyears==mynyears) %>% 
  filter(variable == myvar) %>% 
  filter(species %in% myspecies) %>% 
  filter(as.numeric(substr(assessmentpair,1,4)) >= firstyear) %>%

  ggplot(aes(x=year, y=value, group=assessmentyear)) +
  theme_publication() +
  theme(
    panel.spacing.x = unit(1, "mm"),
    panel.spacing.y = unit(1, "mm"),
    strip.background = element_blank(),
    strip.text.x     = element_text(face="bold", hjust=0, margin = margin(1,0,0,0, "mm")),
    strip.text.y     = element_text(angle = 0),
    axis.text.x      = element_text(angle = 90, vjust = 0.5), 
    legend.position  = "none"
  ) +

  # long black lines
  geom_line(data=d.pairwise.roughly,size=0.5) +

  # short, thick, coloured comparison lines
  geom_line(aes(colour=factor(label)), size=1) +

  # coloured means segments with start and end points
  geom_line(data=d.pairwise.means, aes(x=year, y=value, colour=factor(label)),
            linetype="dashed", lineend="butt", inherit.aes = FALSE) +
  geom_point(data=filter(d.pairwise.means, year==min(year)),
             aes(colour=factor(label)), size=1) +
  geom_point(data=filter(d.pairwise.means, year==max(year)),
             aes(colour=factor(label)), size=1) +

  # phi text
  geom_text (data=d.pcm,
             aes(label = paste0("phi=", round(phi_abs, digits=2))), 
             x=firstplotyear, y=0, hjust=0, vjust=0,  inherit.aes = FALSE) +

  xlim(firstplotyear,lastplotyear) +
  scale_y_continuous(breaks = pretty_breaks(n=3), labels=scientific_format(digits=1)) +
  # scale_x_continuous(breaks = pretty_breaks(n=2)) +
  expand_limits(y=0) +
  labs(x="", y=myvar, title=mytitle) +
  facet_grid(species~assessmentpair, scales = "free_y")

@@

fig_nums(name    = "scalejumps", level = 1, display = FALSE, 
         caption = "Retrospective scale difference (phi) for 9 species derived from historical assessment records. Results from pairwise comparisons over periods of 5 years. Large dots  refer to years when the scale difference is larger than 20%. Colours indicate the decades. Black triangles indicate benchmark assessment years. The bottom part of the panels shows the assessment methods being deployed, the upper part of the panels the stock definitions.")

@@

# threshold <- 0.10
mynyears      <- 5
myremoveyears <- 0
mythreshold   <- 0.2
myvar         <- "stocksize"
mytitle       <- paste0(myvar, ", number of years: ",mynyears, ", threshold: ", 
                        comma_format(digits=1, suffix="%")(100*mythreshold))

@@

tab_nums(name    = "scalejumps2", level = 1, display = FALSE, 
         caption = "Example (mackerel) of sensivity of scale-jumps to the number of years included and the threshold for defining a scale-jump between two assessments.")

@@

myremoveyears <- 0
myvar         <- "stocksize"
myspecies     <- "mac"
mytitle       <- paste0(toupper(myspecies), ", ", myvar, ", removed years: ", myremoveyears)

@@

```{r echo=FALSE, fig.asp=0.8, fig.align="center", message=FALSE, warning=FALSE}

mynyears      <- 5
myremoveyears <- 0
myvar         <- "stocksize"
mytitle       <- paste0(toupper(myspecies), ", ", myvar, ", removed years: ", myremoveyears)

# select combinations of stock and assessmentyear when stock definition has changed
t <-
  d %>% 
  select(assessmentyear, stock, stockkeylabelold) %>% 
  distinct() %>% 
  mutate(
    stock = factor(stock),
    stock = factor(stock, levels=rev(levels(stock)))) %>% 
  group_by(stock, stockkeylabelold) %>% 
  summarize(
    firstyear = min(assessmentyear),
    lastyear  = max(assessmentyear)) %>% 
  arrange(stock, firstyear) %>% 
  group_by(stock) %>% 
  mutate(
    previous       = lag(lastyear),
    assessmentpair = paste(previous, firstyear, sep="-"),
    stockold       = lag(stockkeylabelold)) %>% 
  filter(!is.na(previous))  %>% 
  select(stock, assessmentpair)

# select the assessment information after the change
t1 <-
  t %>% 
  left_join(d.pairs, by=c("stock", "assessmentpair")) %>% 
  select(-assessmentyearprior) %>% 
  left_join(d, by=c("stock", "variable", "assessmentyearlast" = "assessmentyear")) %>% 
  mutate(label="last") %>% 
  rename(assessmentyear = assessmentyearlast)

# data series with previous (prior) estimation
t2 <-
  t %>% 
  left_join(d.pairs, by=c("stock", "assessmentpair")) %>% 
  select(-assessmentyearlast) %>% 
  left_join(d, by=c("stock", "variable", "assessmentyearprior" = "assessmentyear")) %>% 
  mutate(label="prior") %>% 
  rename(assessmentyear = assessmentyearprior)

# combine before and after the change
t3 <-
  bind_rows(t2, t1) %>% 
  group_by(stock, assessmentpair, variable) %>%

  # filter the appropriate period
  filter(year <= max(assessmentyear) - myremoveyears,
         year >= max(assessmentyear) - myremoveyears - mynyears) %>%

# calculate pairwise means
t3.pairwise.means <-

  t3 %>% 
  group_by(header, stock, assessmentpair, variable) %>% 
  distinct(header, stock, assessmentpair, variable) %>% 
  left_join(res$d.pairwise, by=c("stock", "assessmentpair", "variable")) %>%

  filter(
    removelastyears == myremoveyears,
    numberofyears   == mynyears,
    variable        == myvar) %>%

  # calculate means per period
  group_by(header, stock, assessmentpair, variable) %>% 
  mutate(   firstyear      = min(year),
            lastyear       = max(year)) %>% 
  group_by(header, stock, assessmentpair, variable, label, firstyear, lastyear) %>% 
  summarise(value = mean(value, na.rm=TRUE)) %>% 
  gather(key=tmp, value=year, firstyear:lastyear) %>% 
  arrange(header, stock, assessmentpair, variable, label)

# selection of pairwise long timeseries
t3.pairwise.roughly <-

  t3 %>% 
  group_by(header, stock, assessmentpair, variable) %>% 
  distinct(header, stock, assessmentpair, variable) %>% 
  left_join(res$d.pairwise.roughly, by=c("stock", "assessmentpair", "variable")) %>%

  filter(
    removelastyears == myremoveyears,
    numberofyears   == mynyears,
    variable        == myvar)

# get phi for selected assessment pairs
t3.pcm <-

  t3 %>% 
  group_by(header, stock, assessmentpair, variable) %>% 
  distinct(header, stock, assessmentpair, variable) %>% 
  left_join(res$d.pcm, by=c("stock", "assessmentpair", "variable")) %>%

  filter(
    removelastyears == myremoveyears,
    numberofyears   == mynyears,
    variable        == myvar) %>% 
  ungroup() %>% 
  distinct(header, stock, assessmentpair, variable, .keep_all = TRUE)

@@

  filter(variable == myvar) %>%

@@

  # geom_dl(aes(label  = stockkeylabelold, colour = label),
  #         method = list(dl.combine("first.bumpup"), cex = 0.8, hjust=0)) +

  # long black lines
  geom_line(data=t3.pairwise.roughly,size=0.5) +

  # This coloured line segments
  geom_line(aes(colour=label), size=1.5) +

  # coloured means segments with start and end points
  geom_line(data=t3.pairwise.means, aes(x=year, y=value, colour=label),
            linetype="dashed", lineend="butt", inherit.aes = FALSE) +
  geom_point(data=filter(t3.pairwise.means, year==min(year)),
             aes(colour=label), size=1) +
  geom_point(data=filter(t3.pairwise.means, year==max(year)),
             aes(colour=label), size=1) +

  # phi text
  geom_text (data=t3.pcm,
             aes(label = paste0("phi=", round(phi_abs, digits=2)), x= min(t3.pairwise.roughly$year)), 
             y=0, 
             hjust=0, vjust=0,  inherit.aes = FALSE) +

  # annotate("text", label = "left", 
  #           x =  min(t3.pairwise.roughly$year), y = 0, hjust = 0) +

@@

# unique(d.pairwise.means$label)
# unique(d.pairwise.roughly$label)
# unique(t3$label)

@@

```{r, eval=FALSE, fig.align="center", fig.asp=0.5, message=FALSE, warning=FALSE, include=FALSE}

@@

@@

```{r eval=FALSE, fig.align="center", fig.asp=0.7, message=FALSE, warning=FALSE, include=FALSE}

myvar         <- "stocksize"
myspecies     <- "mac"
firstyear     <- 2008
firstplotyear <- 2000
mynyears      <- 5
myremoveyears <- 0
mythreshold   <- 0.1
mytitle       <- paste0(toupper(myspecies), ", ", myvar, ", number of years: ",mynyears, 
                        ", removed years: ", myremoveyears)
